### 2014-01-10 thesis-proposal/problem.tex
Proposal finalized through sec. 1.2

% !TEX root = omar-thesis-proposal.tex
\vspace{-25pt}
\section{Motivation}\label{motivation}
Specifying and implementing a programming language together with its supporting tools (collectively, a \emph{programming system}) that has sound theoretical foundations, helps users identify and fix errors as early as possible, supports a natural programming style, and performs well across diverse problem domains and hardware platforms remains a grand challenge in computing. Due to the increasing diversity and complexity of modern problem domains and hardware platforms, it has become clear that no small  collection of general-purpose constructs backed by a simple, all-purpose implementation can fully satisfy these criteria in all cases. Instead,  researchers and domain experts (collectively, \emph{providers}) continue to develop  specialized notations,  type systems, semantics, implementation strategies, optimizations, run-time systems and tools (collectively, \emph{features})  designed to address these modern challenges in different circumstances. %, researchers and domain experts must continue to design and develop, in a principled manner, specialized notations, type systems, optimizations, run-time systems and tools (collectively, \emph{features}) to support developers as they work in increasingly complex problem domains and on modern hardware systems.

@@

Ideally, a provider would develop and distribute such new features orthogonally, as libraries, so that client developers could granularly choose those that best satisfy their needs. Unfortunately this is often infeasible because libraries are vehicles for specifying a program in terms of a monolithic collection of existing features. From the perspective of a library, the language's syntax, type system and semantics are fixed in advance, the compiler and run-time system are ``black box'' implementations of this fixed language, and the other tools, like code editors and debuggers, operate according to domain-agnostic protocols also based only on a fixed language specification. As a result, providers of new system features must typically take \emph{language-external approaches}, sometimes designing a new language altogether. We will argue that such approaches are problematic, and that taking them has led to an unnecessary gap between research and practice. In place of these approaches, we will advocate for \emph{language-integrated extensibility mechanisms}, and show how, by organizing new features around {types} and imposing certain constraints, these mechanisms can be made both safe and highly expressive. We call a type that introduces new features an \emph{active type} and programming systems organized around active types \emph{actively-typed programming systems}.
%But taking a \emph{language-internal approach} to implementing a feature is the most practical. If a feature can be realized by creatively using existing language constructs and distributed as a library, clients face fewer barriers to adoption because it is easy to integrate library-based features into existing projects gradually and granularly and they leverage well-understood and well-developed mechanisms.
% But taking this approach is often \emph{not} possible today %We call designs \emph{monolithic programming systems}.

@@

%In a conventional \emph{monolithic} programming system, support for each of these features would need to be built into the language and tools. 
No system today has built-in support for all of the features enumerated above. Instead, libraries generally provide support for regular expressions by leveraging  general-purpose constructs. Unfortunately, it is impossible to fully encode the syntax and the specialized static and dynamic semantics described above in terms of general-purpose notations and abstractions. Library providers have thus needed to compromise. The typical choice is to have clients provide regular expressions as strings, deferring parsing, typechecking and elaboration to run-time, which introduces performance overhead and can lead to unanticipated run-time errors (as shown in \cite{spishak2012type}) and security vulnerabilities (due to injection attacks when user inputs are spliced into patterns, for example). Similarly, tools are rarely integrated into editors, and even more rarely in a way that facilitates their discovery and use directly when the developer is manipulating regular expressions. Tools that must be discovered independently and accessed externally (for example, via a browser) are used less frequently  and can be more awkward than necessary \cite{Campbell:2008:DRT:1636642.1636651, Omar:2012:ACC:2337223.2337324}, leading to lower productivity.

\subsection{Language-External Approaches}
\begin{figure}
\begin{center}
\includegraphics[scale=0.45]{approaches.pdf}
\end{center}
\vspace{-20px}
\caption{\small (a) With the language-external approach, novel features are packaged into separate languages and tools. Clients of an API can only safely and naturally use it if the interface uses common constructs (the \emph{compatibility problem}). (b) With the active library approach, there is one extensible host language and the compile-time and edit-time logic governing novel constructs is expressed within libraries. If the extension mechanism guarantees the safety of arbitrary compositions of extensions, the necessary primitives can simply be imported by library clients as needed, and so interoperability is not a problem.}
\label{approaches}
\end{figure}

When the primitive behaviors of a system must be extended to fully realize a new feature, as in the example above, providers typically take a \emph{language-external approach}, either by developing a new or derivative programming system (often centered around a so-called \emph{domain-specific language (DSL)} \cite{fowler2010domain}) or by extending an existing system by some mechanism that is not part of the language itself, such as an extension mechanism for a {particular} compiler\footnote{Compilers that modify, or allow modification of, the semantics of their base language, rather than simply permitting meaning-preserving optimizations, should be considered a pernicious means for creating new languages. Many ``practical'' compilers, including \texttt{gcc}, \texttt{GHC} and \texttt{SML/NJ}, are guilty of this sin, meaning that some programs that seem to be written in C, Haskell or Standard ML are actually written in tool-specific derivatives of those languages. Language-internal mechanisms do not lead to such fragmentation.} or other tool\todo{relate to regex example and update figure}. %This latter method couples the semantics of the feature to the implementation details of a particular tool. Because the use of one implementation entails a different semantics for the feature than another, the extended tool acts, \emph{de facto}, as a distinct system for our purposes.

Unfortunately, when providers of new features take language-external approaches, it causes problems for clients that hinder adoption. Features cannot be adopted individually, but instead are only available coupled to a collection of other unrelated  features (e.g. the incidental features of a newly-designed language or tool, or of the particular tool that was extended). This makes adoption more costly when these incidental features are  not appropriate or insufficiently developed, or when the features bundled with a different language or tool are simultaneously desirable. For example, although evidence suggests that developers prefer language-integrated parallel programming abstractions to library-based implementations if all else is equal \cite{langvslib}, library-based implementations are more widely adopted because ``parallel programming languages'' provide full support only for a few chosen parallel programming abstractions along with a small collection of less-developed general-purpose abstractions. This is problematic because different parallel abstractions are more appropriate in different situations \cite{ecoopstudy13} and parallel programming support is rarely the only concern relevant to client developers outside of a classroom setting. Regular expression support, for example, may be simultaneously desirable when processing large amounts of textual data in parallel. %Similarly, a language and tools designed primarily to support regular expressions might make an interesting research project, but it would not be a suitable tool for writing large applications with more varied needs.

@@

Various concerns can sometimes be separated into different components, each written using a suitable language and tooling. This has been called the \emph{language-oriented approach} to software development \cite{language-oriented} and is supported by \emph{language workbenches} \cite{workbenches} and domain-specific language (DSL) frameworks \cite{fowler10}. Unfortunately, this is insufficient: the interfaces between components written in different languages remains an issue. The specialized constructs particular to one language (e.g. futures in a parallel programming language) cannot always be safely and naturally expressed in terms of those available in another (e.g. a general-purpose language), so building a program out of components written in a variety of different languages is difficult whenever these are exposed at interface boundaries. Tool support is also lost when calling into a different language. We call this fundamental issue the \emph{compatibility problem}.

One strategy often taken by proponents of the language-oriented approach to partially address the compatibility problem is to  target an established intermediate language, such as the Java Virtual Machine (JVM) bytecode\todo{mention functional IL} and support a superset of its constructs. Scala \cite{scala} and F\# \cite{fsharp} are examples of prominent general-purpose languages that have taken this approach, and most DSL frameworks also rely on this strategy. This only enables full interoperability in one direction (Figure \ref{approaches}a). While calling into the common language becomes straightforward, calls in the other direction, or between the languages sharing the common target, are still restricted by the semantics of the common language.

If new languages only include constructs that can already be expressed safely and reasonably naturally in the common language, then this approach can work well. 
But many of the most innovative constructs found in modern languages (often, the features that justify their creation) are difficult to define in terms of existing constructs in ways that guarantee all necessary invariants are statically maintained and that do not require large amounts boilerplate code and run-time overhead. For example, the type system of F\# guarantees that null values cannot occur within F\# data structures, but maintaining this important invariant still requires run-time checks because the typing rules of F\# do not apply when F\# code is called from other languages on the Common Language Infrastructure (CLI) like C\#. The F\# type system also includes support for checking that units of measure are used correctly \cite{fsharpunits}, but this specialized invariant is left completely unchecked at language boundaries. 
%In Scala, traits that have default method implementations are difficult to implement from Java or other JVM languages and the workaround can break if the trait is modified \cite{scalatraitinterop}. 
In some cases, desirable features must be omitted entirely due to concerns about  interoperability. F\#, for example, aims to maintain compatibility with ML, but due to the need for bidirectional interoperability with CLI languages, it cannot support modules or functors.
%\end{itemize}

>>>>

% !TEX root = omar-thesis-proposal.tex
\vspace{-25pt}
\section{Motivation}\label{motivation}
Specifying and implementing a programming language together with its supporting tools (collectively, a \emph{programming system}) that has sound theoretical foundations, helps users identify and fix errors as early as possible, supports a natural programming style, and performs well across diverse problem domains and hardware platforms remains a grand challenge in computing. Due to the increasing diversity and complexity of modern problem domains and hardware platforms, it has become clear that no small  collection of general-purpose constructs backed by a simple, all-purpose implementation can fully satisfy these criteria in all cases. Instead,  researchers and domain experts (collectively, \emph{providers}) continue to develop  specialized notations,  static semantics, dynamic semantics, implementation strategies, optimizations, run-time systems and tools (collectively, \emph{features})  designed to address these modern challenges in different circumstances. %, researchers and domain experts must continue to design and develop, in a principled manner, specialized notations, type systems, optimizations, run-time systems and tools (collectively, \emph{features}) to support developers as they work in increasingly complex problem domains and on modern hardware systems.

@@

Ideally, a provider would develop and distribute such new features orthogonally, as libraries, so that client developers could granularly choose those that best satisfy their needs. Unfortunately this is often infeasible because libraries are vehicles for specifying a program in terms of a monolithic collection of existing features. From the perspective of a library, the language's syntax, type system and semantics are fixed in advance, the compiler and run-time system are ``black box'' implementations of this fixed language, and the other tools, like code editors and debuggers, operate according to domain-agnostic protocols also based only on a fixed language specification. As a result, providers of new system features must often take \emph{language-external approaches}, sometimes designing a new language and its associated tools altogether. We will argue that such approaches are problematic, and that taking them has led to an unnecessary gap between research and practice. In place of these approaches, we will advocate for \emph{language-integrated extensibility mechanisms}, and show how, by organizing new features around {types} enforcing certain constraints, these mechanisms can be made both safe and highly expressive. We call a type that introduces new features an \emph{active type} and programming systems organized around active types \emph{actively-typed programming systems}.
%But taking a \emph{language-internal approach} to implementing a feature is the most practical. If a feature can be realized by creatively using existing language constructs and distributed as a library, clients face fewer barriers to adoption because it is easy to integrate library-based features into existing projects gradually and granularly and they leverage well-understood and well-developed mechanisms.
% But taking this approach is often \emph{not} possible today %We call designs \emph{monolithic programming systems}.

@@

%In a conventional \emph{monolithic} programming system, support for each of these features would need to be built into the language and tools. 
No system today has built-in support for all of the features enumerated above. Instead, libraries generally provide support for regular expressions by leveraging  general-purpose constructs. Unfortunately, it is impossible to fully encode the syntax and the specialized static and dynamic semantics described above in terms of general-purpose notations and abstractions like objects or inductive datatypes. Library providers have thus needed to compromise, typically by asking clients to provide regular expressions as strings, thereby deferring parsing, typechecking and elaboration to run-time, which introduces performance overhead and can lead to unanticipated run-time errors (as shown in \cite{spishak2012type}) and security vulnerabilities (due to injection attacks when user inputs are spliced into patterns, for example). Similarly, useful tools for working with regular expressions are rarely integrated into editors, and even more rarely in a way that facilitates their discovery and use directly when the developer is manipulating regular expressions. Tools that must be discovered independently and accessed externally are used less frequently  and can be more awkward than necessary \cite{Campbell:2008:DRT:1636642.1636651, Omar:2012:ACC:2337223.2337324}, leading to lower productivity \cite{Omar:2012:ACC:2337223.2337324}.

\subsection{Language-External Approaches}
\begin{figure}
\begin{center}
\includegraphics[scale=.48]{approaches.pdf}
\end{center}
\vspace{-20px}
\caption{\small (a) When taking a language-external approach, new features are packaged together into separate languages and tools, causing problems with compatibility (described in the text). (b) When taking a language-integrated approach, there is one extensible host language and the compile-time and edit-time logic governing new constructs is expressed within ``active'' libraries. It is critical that the extension mechanism  guarantees that extensions cannot interfere with one another, so that the compatibility problem is avoided.}
\label{approaches}
\end{figure}

When the semantics or implementation of a system must be extended to fully realize a new feature, as in the example above, providers typically take a \emph{language-external approach}, either by developing a new or derivative programming system (supported by \emph{language workbenches} \cite{erdweg2013state},  \emph{DSL frameworks} \cite{fowler2010domain} or \emph{compiler generators}), or by extending an existing system by some mechanism that is not part of the language itself, such as an extension mechanism for a {particular} compiler\footnote{Compilers that modify, or allow modification of, the semantics of their base language, rather than simply permitting meaning-preserving optimizations, should be considered a pernicious means for creating new languages. Many ``practical'' compilers, including \texttt{gcc}, \texttt{GHC} and \texttt{SML/NJ}, are guilty of this sin, meaning that some programs that seem to be written in C, Haskell or Standard ML are actually written in tool-specific derivatives of those languages. Language-internal mechanisms do not lead to such fragmentation.} or other tool. For example, a researcher interested in providing regexp-related features (let us refer to these collectively as \texttt{R}) might design a new language with built-in support for regular expressions, perhaps basing it on an existing language (containing some general-purpose constructs, \texttt{G}). A different researcher developing a new language-based parallel programming abstraction or implementation strategy (collectively, \texttt{P}) might  take the same strategy. A third researcher, developing an alternative parallel programming abstraction (collectively, \texttt{Q}) might again do the same. This results in a collection of distinct systems as diagrammed in Figure \ref{approaches}a.%This latter method couples the semantics of the feature to the implementation details of a particular tool. Because the use of one implementation entails a different semantics for the feature than another, the extended tool acts, \emph{de facto}, as a distinct system for our purposes.

Unfortunately, when providers of new features take language-external approaches like this, it causes  problems for clients. Features cannot be adopted individually, but instead are only available coupled to a collection of other unrelated  features (e.g. the incidental features of the newly-designed language and its associated tools). This makes adoption more costly when these incidental features are  not desirable or insufficiently developed, or when the features bundled with a different language or tool are simultaneously desirable. Evidence supports the prominence of this problem: developers prefer language-integrated parallel programming abstractions to library-based implementations if all else is equal \cite{cave2010comparing}, but library-based implementations are more widely adopted because ``parallel programming languages'' privilege only a few chosen  abstractions and implementation strategies. This is problematic because different parallel abstractions or implementation strategies are more appropriate in different situations \cite{Tasharofi:2013rc} and moreover,  parallel programming support is rarely the only concern relevant to client developers outside of a classroom setting. Regular expression support, for example, may be simultaneously desirable for processing large amounts of textual data in parallel, but using these features together in the same compilation unit when they have been implemented by language-external means is difficult or impossible. One must either use the system containing features \texttt{G+R} or \texttt{G+P}. There is no system containing both. %Similarly, a language and tools designed primarily to support regular expressions might make an interesting research project, but it would not be a suitable tool for writing large applications with more varied needs.

@@

Different application concerns can sometimes be separated into different components, each written using a suitable language and tooling. This has been called the \emph{language-oriented approach} to software development \cite{journals/stp/Ward94}. Unfortunately, this is insufficient: the interface between components written in different languages remains an issue. The specialized constructs particular to one language (e.g. futures in a parallel programming language) cannot always be safely and naturally expressed in terms of those available in another (e.g. a general-purpose language), so building a program out of components written in a variety of different languages is difficult whenever these are exposed at interface boundaries. Tool support is also lost when calling into a different language. We call this fundamental issue the \emph{client compatibility problem}: clients of a certain collection of features cannot always interface with clients of a different collection of features in a safe, performant and natural manner.

One strategy often taken by proponents of the language-oriented approach to partially address the compatibility problem is to  target an established intermediate language, such as the Java Virtual Machine (JVM) bytecode, and support a superset of its constructs. Scala \cite{200464/IC} and F\# \cite{pickering2007foundations} are examples of prominent general-purpose languages that have taken this approach, and most DSL frameworks also rely on this strategy. As indicated in Figure \ref{approaches}a, this only enables true client compatibility in one direction. While calling into the common language becomes straightforward, calls in the other direction, or between the languages sharing the common target, are still restricted by the semantics of the common language.

If new languages only include constructs that can already be expressed safely and reasonably naturally in the common language, then this approach can work well.
But many of the most innovative constructs found in modern languages (often, the features that justify their creation) are difficult to define in terms of existing constructs in ways that guarantee all necessary invariants are statically maintained and that do not require large amounts boilerplate code and run-time overhead. For example, the type system of F\# guarantees that null values cannot occur within F\# data structures, but maintaining this important invariant still requires run-time checks because the typing rules of F\# do not apply when F\# code is called from other languages on the Common Language Infrastructure (CLI) like C\#. The F\# type system also includes support for checking that units of measure are used correctly \cite{syme2012expert, kennedy1994dimension}, but this specialized invariant is left completely unchecked at language boundaries. 
%In Scala, traits that have default method implementations are difficult to implement from Java or other JVM languages and the workaround can break if the trait is modified \cite{scalatraitinterop}. 
In some cases, desirable features must be omitted entirely due to concerns about  interoperability. F\#, for example, is based on Ocaml, but due to the need for bidirectional interoperability with CLI languages, it cannot support advanced features like polymorphic variants, modules or functors \cite{ocaml-manual}.
%\end{itemize}
